---
title: "Applied example: daily LPA on the minute / time-ratio scale"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Applied example: daily LPA on the minute / time-ratio scale}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>",
                      fig.width = 6.5, fig.height = 4.2)
```

The quickstart reports the dose-response on the internal **log** scale — the
analysis axis of the AFT model, `log T = b0 + f(log W) + ...`. For an applied
audience the same fit is more interpretable on two transformed axes:

- **x-axis in minutes.** Exponentiate the log-exposure grid, `exp(x_grid)`, so
  the curve is read against daily LPA in minutes rather than `log` units.
- **y-axis as a time-acceleration ratio.** Exponentiate the centred curve,
  `TR(w) = exp{f(log w)}`. Under the AFT model a `TR` of 1.2 means survival
  time is accelerated (delayed) by 20% relative to the reference exposure.

This article fits the pipeline on simulated daily-LPA data and presents it the
way the manuscript reports the applied result: **time ratio vs minutes**,
anchored at a clinical reference of **200 min/day**, restricted to the
interpretable exposure window.

```{r setup}
library(aftsplex)
have_gg <- requireNamespace("ggplot2", quietly = TRUE)
```

## Simulate on the daily-LPA log scale

```{r sim}
XM <- log(300); XS <- 0.45                 # ~300 min/day, 95% ~ [124, 725] min
FA <- list(x_min = log(150), alpha = 1.5, beta = 0.6)
V_REF <- c(V1 = 30, V2 = 30, V3 = 0, V4 = 0)

set.seed(2026)
Su  <- sigma_u_for_reliability(0.7, var_x = XS^2)   # mean reliability 0.70
sim <- generate_aft_data(n = 1000, n_val = 300, Sigma_u = Su,
                         x_mean = XM, x_sd = XS, f_args = FA)
cal <- fit_me_calibration(sim$validation)
g   <- gls_combine(as.matrix(sim$survival[, cal$W_cols]), cal)
dat <- sim$survival; dat$W_bar <- g$W_bar
```

## Anchor and exposure window

Two applied choices matter here. First, the reference: instead of anchoring the
curve at the sparse left edge of the grid (where the spline is least supported
and the band is widest), we anchor `TR = 1` at a clinically meaningful
**200 min/day** via `x_ref`. Second, the window: the natural cubic spline goes
linear beyond the boundary knots and the data thin below ~150 min, so we report
only over the interpretable range (here, `>= 150 min` up to the 95th
percentile) rather than extrapolating into the sparse tail.

```{r grid}
REF_MIN <- 200; LOW_MIN <- 150
x_lo   <- max(log(LOW_MIN), quantile(dat$W_bar, 0.05))
x_hi   <- quantile(dat$W_bar, 0.95)
x_grid <- seq(x_lo, x_hi, length.out = 40)
x_ref  <- log(REF_MIN)
LAMBDA <- c(0.5, 1, 1.5, 2)

s <- simex_aft_spline(dat, x_var = "W_bar", sigma_w_sq = g$sigma_w_sq,
                      covariates = names(V_REF), v_ref = V_REF,
                      lambda = LAMBDA, B = 30, x_grid = x_grid, x_ref = x_ref)

boot <- two_stage_bootstrap(
  survival = sim$survival, validation = sim$validation, x_var = "X_true",
  covariates = names(V_REF), v_ref = V_REF, lambda = LAMBDA,
  B = 25, R = 40, x_grid = x_grid, x_ref = x_ref, verbose = FALSE)
```

## Headline figure: time ratio vs LPA minutes

The corrected (SIMEX) curve sits above the uncorrected (naive) curve at high
LPA: ignoring measurement error attenuates the dose-response toward the null, so
the naive time ratio understates the benefit. Both are anchored at `TR = 1` at
200 min/day; the shaded band is the two-stage bootstrap 95% interval.

```{r headline, eval = have_gg, fig.width = 7, fig.height = 4.5}
library(ggplot2)
df <- data.frame(min = exp(x_grid),
                 Naive = exp(s$curve_naive), SIMEX = exp(s$curve_simex),
                 lower = exp(boot$lower), upper = exp(boot$upper))
df_long <- rbind(
  data.frame(min = df$min, tr = df$Naive, method = "Naive"),
  data.frame(min = df$min, tr = df$SIMEX, method = "SIMEX"))
cols <- method_colors()[c("Naive", "SIMEX")]

ggplot() +
  geom_hline(yintercept = 1, color = "grey60", linewidth = 0.4) +
  geom_vline(xintercept = REF_MIN, color = "grey40",
             linetype = "dashed", linewidth = 0.5) +
  geom_ribbon(data = df, aes(min, ymin = lower, ymax = upper),
              fill = cols[["SIMEX"]], alpha = 0.22) +
  geom_line(data = df_long, aes(min, tr, color = method), linewidth = 1) +
  scale_color_manual(values = cols, name = NULL,
                     labels = c(Naive = "Uncorrected (naive)",
                                SIMEX = "SIMEX-corrected")) +
  labs(x = "Light physical activity (minutes/day)",
       y = "Time-acceleration ratio (relative to 200 min/day)") +
  theme_bw(base_size = 12) +
  theme(legend.position = c(0.99, 0.02), legend.justification = c(1, 0))
```

## The untransformed curve (appendix view)

The same fit on the internal log scale — the centred linear predictor
`f(log w)` — is what the simulation appendix reports. The applied figure above
is just this curve with both axes exponentiated.

```{r untransformed, eval = have_gg}
f_d   <- function(x) do.call(f_true, c(list(x), FA))
truth <- f_d(x_grid) - f_d(x_grid[1])
plot(s, truth = truth, title = "Centred log-time-ratio (analysis scale)")
```

## How much does Phase-1 uncertainty add?

The two-stage (`nested = TRUE`) bootstrap resamples both the validation sample
and the main study, so the band carries Phase-1 calibration uncertainty. Holding
Phase-1 fixed (`nested = FALSE`) isolates the main-study contribution; the ratio
of mean widths quantifies how much the finite validation sample inflates the
interval.

```{r nested}
boot_single <- two_stage_bootstrap(
  survival = sim$survival, validation = sim$validation, x_var = "X_true",
  covariates = names(V_REF), v_ref = V_REF, lambda = LAMBDA,
  B = 25, R = 40, x_grid = x_grid, x_ref = x_ref, nested = FALSE)

w_two    <- mean(boot$upper - boot$lower, na.rm = TRUE)
w_single <- mean(boot_single$upper - boot_single$lower, na.rm = TRUE)
c(two_stage = w_two, single_stage = w_single,
  inflation = w_two / w_single)
```

When the validation sample is small relative to the main study, the two-stage
width is appreciably larger — which is exactly why the nested design is the
default for reported intervals.

## Parallel execution

The outer bootstrap loop parallelises over `workers`; see `vignette("faq")` for
the reproducibility caveat (the serial `workers = 1` path is the canonical
bit-for-bit reproducible one).

```{r workers, eval = FALSE}
two_stage_bootstrap(sim$survival, sim$validation, x_var = "X_true",
                    covariates = names(V_REF), v_ref = V_REF,
                    x_grid = x_grid, x_ref = x_ref, R = 200, workers = 4)
```
