---
title: "Quick start: AFT + spline + SIMEX in 30 lines"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Quick start: AFT + spline + SIMEX in 30 lines}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width = 6, fig.height = 4
)
```

This vignette walks through one end-to-end fit on simulated data, taking
under ten seconds on a laptop. The intended workflow on real data is the
same: a small-validation Phase-1 fit, GLS-combined exposure on the main
study, SIMEX correction, and a two-stage bootstrap for inference.

```{r setup}
library(aftsplex)
```

## Simulate

`generate_aft_data()` returns a list with a main-study survival data frame,
a validation data frame, and the truth used to generate them.

```{r sim}
sim <- generate_aft_data(n = 1000, n_val = 300, seed = 1)
head(sim$survival, 3)
```

## Phase-1 calibration

```{r phase1}
cal <- fit_me_calibration(sim$validation)
cal$alpha1                # close to 1: surrogate is unbiased for truth
round(cal$Sigma_e, 2)
```

`gls_combine()` collapses the three correlated surrogates into a single
calibrated exposure with conditional variance `sigma_w_sq`.

```{r gls}
W <- as.matrix(sim$survival[, cal$W_cols])
g <- gls_combine(W, cal)
dat <- sim$survival
dat$W_bar <- g$W_bar
round(g$sigma_w_sq, 3)
```

## SIMEX point estimate

```{r simex}
x_grid <- seq(quantile(g$W_bar, 0.02),
              quantile(g$W_bar, 0.98), length.out = 50)

s <- simex_aft_spline(
  dat, x_var = "W_bar", sigma_w_sq = g$sigma_w_sq,
  covariates = c("V1","V2","V3","V4"),
  v_ref = c(V1 = 30, V2 = 30, V3 = 0, V4 = 0),
  lambda = c(0.5, 1, 1.5, 2), B = 10,
  x_grid = x_grid
)
```

## Two-stage bootstrap interval

`two_stage_bootstrap()` runs the SIMEX pipeline inside an outer loop that
resamples both the validation and the main-study data, so its interval
propagates Phase-1 and Phase-2 sampling variation jointly.

```{r boot}
set.seed(1)
boot <- two_stage_bootstrap(
  survival   = sim$survival,
  validation = sim$validation,
  x_var      = "X_true",                # truth column for calibration
  covariates = c("V1","V2","V3","V4"),
  v_ref      = c(V1 = 30, V2 = 30, V3 = 0, V4 = 0),
  lambda     = c(0.5, 1, 1.5, 2), B = 10, R = 30,
  x_grid     = x_grid,
  verbose    = FALSE
)
```

## Plot

```{r plot, eval = requireNamespace("ggplot2", quietly = TRUE)}
truth <- f_true(x_grid) - f_true(x_grid[1])
fits  <- list(SIMEX = boot$f_hat)
plot_curves(x_grid, truth, fits,
            ci = list(lower = boot$lower, upper = boot$upper),
            title = "SIMEX-corrected dose-response with two-stage bootstrap CI")
```
