---
title: "Simulation study: Oracle vs Naive vs SIMEX"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Simulation study: Oracle vs Naive vs SIMEX}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>",
                      fig.width = 7, fig.height = 4)
```

A small Monte Carlo replicating the structure of the manuscript appendix
simulation. Three estimators are compared on the same `r` replicates:

- **Oracle** -- fits on the true latent exposure `X_true` (an upper bound).
- **Naive** -- fits on the GLS-combined surrogate `W_bar` without ME correction.
- **SIMEX** -- fits on `W_bar` with the SIMEX correction.

The vignette uses `N_REP = 20` replicates so it builds in under a minute;
the manuscript figure uses 500. Set `N_REP` higher locally to reproduce.

```{r setup}
library(aftsplex)

N_REP   <- 20L
N       <- 1500
N_VAL   <- 400
B_INNER <- 10
DF      <- 4
LAMBDA  <- c(0.5, 1, 1.5, 2)

x_grid <- seq(qnorm(0.02, 10, sqrt(5)),
              qnorm(0.98, 10, sqrt(5)),
              length.out = 100)
truth  <- f_true(x_grid) - f_true(x_grid[1])
V_REF  <- c(V1 = 30, V2 = 30, V3 = 0, V4 = 0)
```

## Single-replicate driver

```{r driver}
run_one <- function(seed) {
  sim <- generate_aft_data(n = N, n_val = N_VAL, seed = seed)
  cal <- fit_me_calibration(sim$validation)
  W   <- as.matrix(sim$survival[, cal$W_cols])
  g   <- gls_combine(W, cal)
  dat <- sim$survival
  dat$W_bar <- g$W_bar

  fit_oracle <- fit_aft_spline(dat, "X_true",
                               covariates = names(V_REF), df = DF)
  fit_naive  <- fit_aft_spline(dat, "W_bar",
                               covariates = names(V_REF), df = DF)
  lp_oracle <- predict_curve(fit_oracle, x_grid, V_REF, "X_true")
  lp_naive  <- predict_curve(fit_naive,  x_grid, V_REF, "W_bar")

  s <- simex_aft_spline(dat, x_var = "W_bar",
                        sigma_w_sq = g$sigma_w_sq,
                        covariates = names(V_REF), v_ref = V_REF,
                        lambda = LAMBDA, B = B_INNER,
                        x_grid = x_grid)
  list(Oracle = lp_oracle - lp_oracle[1],
       Naive  = lp_naive  - lp_naive[1],
       SIMEX  = s$curve_simex)
}
```

## Run the Monte Carlo

```{r run, results = "hide"}
t0 <- Sys.time()
res <- lapply(seq_len(N_REP), function(i) run_one(20260601L + i))
elapsed <- Sys.time() - t0
```

```{r elapsed}
elapsed
```

## ISE summary

```{r ise}
methods <- c("Oracle", "Naive", "SIMEX")
ise_mat <- sapply(res, function(r)
  vapply(methods, function(m) {
    e <- r[[m]] - truth
    delta_x <- diff(x_grid)
    sum(((e[-1] + e[-length(e)])^2 / 4) * delta_x)
  }, numeric(1))
)
rownames(ise_mat) <- methods
data.frame(
  Mean   = round(rowMeans(ise_mat), 3),
  MCSE   = round(apply(ise_mat, 1, sd) / sqrt(N_REP), 4),
  Median = round(apply(ise_mat, 1, median), 3)
)
```

The pattern matches the manuscript: Naive carries 2-3x the ISE of Oracle
(attenuation), and SIMEX recovers most of that gap.

## Mean curves

```{r curves, eval = requireNamespace("ggplot2", quietly = TRUE)}
mean_curve <- function(m) rowMeans(sapply(res, function(r) r[[m]]))
fits <- setNames(lapply(methods, mean_curve), methods)
plot_curves(x_grid, truth, fits,
            title = "Mean SIMEX / Naive / Oracle curves across simulated replicates")
```

# A daily-minutes scale (mean ~300 min)

The arbitrary-unit scale above keeps the code minimal, but the manuscript
expresses the exposure as **daily light-physical-activity minutes**. The DGP
arguments `x_mean`, `x_sd`, and `f_args` re-express the same model on that
scale -- mean ~300 min, a lower threshold of 150 min below which the
dose-response is flat -- without touching the package defaults.

```{r minutes-setup}
SC <- list(x_mean = 300, x_sd = 75,
           f_args = list(x_min = 150, alpha = 0.01, beta = 0.6))

f_min     <- function(x) do.call(f_true, c(list(x), SC$f_args))
x_grid_m  <- seq(qnorm(0.05, SC$x_mean, SC$x_sd),
                 qnorm(0.95, SC$x_mean, SC$x_sd), length.out = 80)
truth_m   <- f_min(x_grid_m) - f_min(x_grid_m[1])

N_REP_M <- 10L        # small for a fast build; raise locally for the manuscript
N_M     <- 1200
N_VAL_M <- 350

ise_of <- function(curve) {
  e  <- curve - truth_m
  dx <- diff(x_grid_m)
  sum(((e[-1] + e[-length(e)])^2 / 4) * dx)
}
```

## Reliability sweep

How well does SIMEX recover the curve as the surrogate gets noisier? We sweep
the **mean per-surrogate reliability** over `0.5, 0.65, 0.8` (poor -> good),
holding everything else fixed; `sigma_u_for_reliability()` returns the matching
measurement-error covariance.

```{r rel-driver}
run_full <- function(seed, Sigma_u) {
  sim <- generate_aft_data(n = N_M, n_val = N_VAL_M, Sigma_u = Sigma_u,
                           x_mean = SC$x_mean, x_sd = SC$x_sd,
                           f_args = SC$f_args, seed = seed)
  cal <- fit_me_calibration(sim$validation)
  W   <- as.matrix(sim$survival[, cal$W_cols])
  g   <- gls_combine(W, cal)
  dat <- sim$survival; dat$W_bar <- g$W_bar

  fo <- fit_aft_spline(dat, "X_true", covariates = names(V_REF), df = DF)
  fn <- fit_aft_spline(dat, "W_bar",  covariates = names(V_REF), df = DF)
  lo <- predict_curve(fo, x_grid_m, V_REF, "X_true")
  ln <- predict_curve(fn, x_grid_m, V_REF, "W_bar")
  s  <- simex_aft_spline(dat, "W_bar", sigma_w_sq = g$sigma_w_sq,
                         covariates = names(V_REF), v_ref = V_REF,
                         lambda = LAMBDA, B = B_INNER, x_grid = x_grid_m)
  list(Oracle = lo - lo[1], Naive = ln - ln[1], SIMEX = s$curve_simex)
}

rel_levels <- c(0.5, 0.65, 0.8)
sweep <- t(sapply(rel_levels, function(r) {
  Su  <- sigma_u_for_reliability(r, var_x = SC$x_sd^2)
  rep <- lapply(seq_len(N_REP_M), function(i) run_full(70000L + i, Su))
  vapply(methods, function(m) mean(vapply(rep, function(z) ise_of(z[[m]]), 0)),
         numeric(1))
}))
rownames(sweep) <- sprintf("reliability = %.2f", rel_levels)
round(sweep, 3)
```

As reliability falls, the **Naive** ISE inflates sharply from worsening
attenuation, and **SIMEX** removes a large part of that excess -- roughly
halving the Naive ISE at reliability 0.5. The gain shrinks as reliability
improves: by 0.8 the naive attenuation is already small, so the correction has
little left to do (and its extrapolation variance can offset the tiny remaining
bias). SIMEX earns its keep most when the measurement is poorest.

## Per-surrogate ladder

The GLS combiner pools the three surrogates. Here we give the surrogates
distinctly different quality (relative error variances `1 : 3 : 6`, a mean
reliability of 0.65 spanning roughly 0.84 down to 0.47) and run the estimator on
each surrogate **alone** (`surrogate_pattern = "^W1$"`, etc.) and on the
combiner. This uses the existing `surrogate_pattern` argument -- no engine
change.

```{r ladder}
patterns <- c(W1 = "^W1$", W2 = "^W2$", W3 = "^W3$", Combined = "^W[0-9]+$")
Su_mid   <- sigma_u_for_reliability(0.65, var_x = SC$x_sd^2, ratios = c(1, 3, 6))

ladder_curves <- lapply(seq_len(N_REP_M), function(i) {
  sim <- generate_aft_data(n = N_M, n_val = N_VAL_M, Sigma_u = Su_mid,
                           x_mean = SC$x_mean, x_sd = SC$x_sd,
                           f_args = SC$f_args, seed = 80000L + i)
  sapply(patterns, function(p) {
    cal <- fit_me_calibration(sim$validation, surrogate_pattern = p)
    W   <- as.matrix(sim$survival[, cal$W_cols, drop = FALSE])
    g   <- gls_combine(W, cal)
    dat <- sim$survival; dat$W_bar <- g$W_bar
    s   <- simex_aft_spline(dat, "W_bar", sigma_w_sq = g$sigma_w_sq,
                            covariates = names(V_REF), v_ref = V_REF,
                            lambda = LAMBDA, B = B_INNER, x_grid = x_grid_m)
    s$curve_simex
  })
})

ladder_ise <- rowMeans(sapply(ladder_curves,
                              function(M) apply(M, 2, ise_of)))
data.frame(SIMEX_ISE = round(ladder_ise, 3))
```

The single-surrogate ISE ranks by reliability (`W1` best, `W3` noisiest), and
the combiner sits at or below even the best single channel -- the payoff for
pooling correlated measurements.

```{r ladder-plot, eval = requireNamespace("ggplot2", quietly = TRUE)}
mean_ladder <- sapply(names(patterns), function(nm)
  rowMeans(sapply(ladder_curves, function(M) M[, nm])))

pal <- stagill_palette()
ladder_cols <- c(W1 = unname(pal["MutedTeal"]),
                 W2 = unname(pal["ApricotCream"]),
                 W3 = unname(pal["BurntPeach"]),
                 Combined = unname(pal["TwilightIndigo"]))

plot_curves(x_grid_m, truth_m,
            fits = as.list(as.data.frame(mean_ladder)),
            colors = ladder_cols,
            x_lab = "Daily LPA (minutes)",
            title = "Per-surrogate ladder: single channels vs the GLS combiner")
```
